Analysing Constraint Grammar with a SAT solver

### 2015-10-15 paper/lrec/problem_desc.tex
more stuff

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}
\end{itemize}
the rule will apply to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
However, if a word has only one remaining analysis, then the rule will not apply, even if the condition is met.
%Rules are not allowed to remove the last remaining analysis: if  \texttt{verb pl} were the only analysis for \emph{bear}, then the rule would not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser resourced languages. They are robust and can be written without large corpora--only morphological analysis is needed. The formalism is lightweight and language-independent, and resources can be shared between related languages \cite{bick2006spanish,lene_trond_linda2010}.
Mature CGs contain some thousands of rules, but even small CGs are shown to be effective \cite{lene_trond2011}.

@@

\begin{itemize}
\item Does my grammar contain rules that contradict each other?
\item Does my grammar contain rules that will never fire?
\item Generate a sequence that triggers rule(s) X but not rule(s) Y
\item Generate a sequence that is ambiguous but doesn't trigger any rules
\end{itemize}

The technique requires an existing morphological lexicon, but no corpus. 
The lexicon is needed in order to generate sequences that have realistic ambiguities: e.g. a word can be ambiguous between noun and verb, but not between adjective and punctuation. 
Constraints for the structure of the sequences come from the CG rules themselves.
We encode both of these constraints as a satisfiability problem, and use a SAT solver to generate answers to the previous questions.

The paper is structured as follows. Section~\ref{sec:prev} relates our work to the previous research. Section~\ref{sec:implementation} discusses the techniques we use, Section~X presents preliminary results and  Section~Y discusses future work. Section~Z concludes the paper.

\todo{Relate to conf themes: "Methodologies and tools for LRs construction and annotation" and
"Validation and quality assurance of LRs"}

\section{Previous work}
\label{sec:prev}

\begin{quote}Another desirable facility in the grammar development environment would
be a mechanism for identifying pairs of constraints that contradict each
other.
--Atro Voutilainen, 2004
\end{quote}

This section describes prior research and relates our contribution to the existing approaches.
We combine elements from the following aspects of CG research:

@@

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}. For a language with no free or tagset-compatible corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to unannotated Wikipedia texts and pick 100 examples at random for manual check.

% CG rules are usually arranged in sections, from safest to least safe. Given a grammar that has three sections, the rules are applied to the text in the following manner. First apply rules from section 1, and repeat until nothing changes in the text. Then apply rules from sections 1--2, and finally all rules, from sections 1--3.
% The best strategy is to start with the most effective rules with low error rate, so that they make way for the following, more heuristic and less safe rules to act on.

A representative corpus is arguably the best way to get concrete numbers: how many times a rule applied and how often it was correct.
However, this method will not notice a missed opportunity, nor suggest a way to improve, for instance changing the place of a rule.

@@

\paragraph{Optimising hand-written grammars}
\begin{itemize}
\item Variations in the place of the rule
\item Variations in the conditions (careful, ...) of the rule
\end{itemize}
Good side: try all combinations to find the best result. 
At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.

@@

\cite{lager98} presents a CG-like shallow parsing system encoded in logic, and \cite{lager_nivre01} continues with a reconstruction of four different formalisms.
%The earlier works on logical reconstruction don't envision grammar analysis as one of the use cases,
Grammar analysis is a natural use case, due to some key features of the logical reconstruction.

The/A(n) 
\todo{imperative/straight-forward/efficient/...?}
CG compiler
%, such as VISL CG-3,
 cannot capture any dependencies between rules.
% It discards analyses immediately: the output of the $i^{th}$ rule becomes the input of the $i+1^{th}$ rule, but there is no information which rules have been applied before.
In contrast, a logic-based CG compiler does that by default. 
The rules are modelled as implications and composed in the order of the rule sequence, such that 
the consequent from the $i^{th}$ rule becomes the antecedent of the $i+1^{th}$ rule.
Given this design, we added on top a way to ask for solutions with certain properties.
%, such as ``after applying rules $0-i$, is it possible for $i+1$ to apply''.

>>>>

Constraint Grammar (CG, \cite{karlsson1995constraint})
% \citep[CG][]{karlsson1995constraint} 
is a method for disambiguating morphologically analysed text% and adding syntactic or dependency labels
. 
A grammar consists of rules that target specific analyses for selection or removal, based on contextual tests. For example, the following rule
\begin{itemize}
\item[] \texttt{REMOVE verb IF (-1C det) ;}
\end{itemize}
removes a verb reading from a word which is preceded by an unambiguously (C) tagged determiner.
Given the following text,
\begin{itemize}
\item[] 
\begin{verbatim}
"<the>"
        "the" det def
"<bear>"
        "bear" noun sg
        "bear" verb pl
\end{verbatim}
\end{itemize}
the rule will match to the word \emph{bear}, and remove the analysis \texttt{verb pl}.
However, if the target word has only one remaining analysis, then the rule will not apply, even if the condition is met.

CGs are valuable resources for rule-based NLP, especially for lesser resourced languages. They are robust and can be written without large corpora--only morphological analysis is needed. The formalism is lightweight and language-independent, and resources can be shared between related languages \cite{bick2006spanish,lene_trond_linda2010}.
%Mature CGs contain some thousands of rules, but even small CGs are shown to be effective \cite{lene_trond2011}.

@@

\begin{itemize}
\item Are there rules that contradict each other?
\item Are there rules that will never fire?
\item Generate a sequence that triggers rule(s) X but not rule(s) Y
\item Generate a sequence that is ambiguous but doesn't trigger any rules
\end{itemize}

This technique requires an existing morphological lexicon, but no corpus. 
The lexicon is needed in order to generate realistic ambiguities: e.g. a word can be ambiguous between noun and verb, but not between adjective and punctuation. 
Constraints for the structure of the sequences come from the CG rules themselves.
We encode both of these constraints as a satisfiability problem, and use a SAT solver to generate answers to the previous questions.

The paper is structured as follows. Section~\ref{sec:prev} relates our work to the previous research. Section~\ref{sec:implementation} discusses the implementation, Section~X presents preliminary results and  Section~Y discusses future work. Section~Z concludes the paper.

%\todo{Relate to conf themes: "Methodologies and tools for LRs construction and annotation" and
%"Validation and quality assurance of LRs"}

\section{Related work}
\label{sec:prev}

% \begin{quote}Another desirable facility in the grammar development environment would
% be a mechanism for identifying pairs of constraints that contradict each
% other.
% --Atro Voutilainen, 2004
% \end{quote}

%This section describes prior research and relates our contribution to the existing approaches.
We combine elements from the following aspects of CG research:

@@

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}.
%\cite{voutilainen2004} gives a detailed account about best practices of grammar writing and efficient use of corpora to aid the grammar development.
%For a language with no free or tagset-compatible corpus available, \cite{tyers_reynolds2015} describe a method where they apply their rules to unannotated Wikipedia texts and pick 100 examples at random for manual check.

CG rules are usually arranged in sections, and run in the following manner. 
First apply rules from section 1, and repeat until nothing changes in the text. Then apply rules from sections 1--2, then 1--3 and so on, until the set includes all rules.
The best strategy is to place the safest and most effective rules in the first sections,
so that they make way for the following, more heuristic and less safe rules to act on.

A representative corpus is arguably the best way to get concrete numbers---how many times a rule applied and how often it was correct---and to arrange the rules in sections based on that feedback.
However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.
%, for instance changing the place of a rule.

@@

\paragraph{Optimising hand-written grammars}
The corpus-based method can tell the effect of each single rule at their place in the rule sequence, and it is up to the grammar writer to make changes in the grammar.
\cite{bick2013tuning} automates the latter process: the rules are moved into different sections or removed in total based on their effect, and in parallel, their contexts are made stricter or less strict.
After trying out several combinations, the version that produces the best results is kept.

This is a valuable tool for handling grammars that are too big to keep track of, and Bick's evaluation shows that the tuned grammars perform better than the original.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.
% At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
% This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.

@@

%\cite{lager98} presents a CG-like shallow parsing system encoded in logic, and \cite{lager_nivre01} continues with a reconstruction of four different formalisms.
%The earlier works on logical reconstruction don't envision grammar analysis as one of the use cases,
\cite{lager98} and \cite{lager_nivre01} reconstruct the CG formalism in first-order predicate logic, and \cite{listenmaa_claessen2015} implement a CG compiler using a SAT solver.
Grammar analysis is a natural use case, due to some key features of the logical reconstruction.
The traditional CG compiler 
%, such as VISL CG-3,
 cannot capture any dependencies between rules.
% It discards analyses immediately: the output of the $i^{th}$ rule becomes the input of the $i+1^{th}$ rule, but there is no information which rules have been applied before.
In contrast, a logic-based CG compiler does that by default. 
The rules are modelled as implications and composed in the order of the rule sequence, such that 
the consequent from the $i^{th}$ rule becomes the antecedent of the $i+1^{th}$ rule.
Given this design, we added on top a way to ask for solutions with certain properties.
%, such as ``after applying rules $0-i$, is it possible for $i+1$ to apply''.
